---
title: "Anemia prevalence among women: an `rdhs` example"
author: "OJ Watson, Jeff Eaton"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Anemia prevalence among women: an `rdhs` example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Anemia is a common cause of fatigue, and women of childbearing age
are at particularly high risk for anemia. The package `rdhs` can be used
to compare estimates of the prevalence of any anemia among women from
Demographic and Health Surveys (DHS) conducted in Armenia, Cambodia,
and Lesotho.

# Setup

Load the `rdhs` package and other useful packages for analysing data.

```{r, results=FALSE, message = FALSE, warning=FALSE}
devtools::install_github("OJWatson/rdhs")
library(rdhs)
library(data.table)
library(ggplot2)
library(survey)
library(haven)
```

# Using calculated indicators from STATcompiler

Anemia prevalence among women is reported as a core indicator through the DHS STATcompiler (https://www.statcompiler.com/).
These indicators can be accessed directly from R via the DHS API with the function `dhs_data()`.

Query the API for a list of all StatCompiler indicators, and then search the indicators for those
that have `"anemia"` in the indicator name. API calls return `data.table` objects. If package `data.table`
is not loaded, they will be printed as `data.frame`s instead.

```{r }
library(rdhs)

indicators <- dhs_indicators()
tail(indicators[grepl("anemia", Label), .(IndicatorId, ShortName, Label)])
```

The indicator ID `"AN_ANEM_W_ANY"` reports the percentage of women with any anemia.
The function `dhs_data()` will query the indicator dataset for the value of this indicator
for our three countries of interest. First, use `dhs_countries()` to query the
list of DHS countries to identify the DHS country code for each country.

```{r }
countries <- dhs_countries()
dhscc <- countries[CountryName %in% c("Armenia", "Cambodia", "Lesotho"), DHS_CountryCode]
dhscc
```

Now query the indicators dataset for the women with any anemia indicator for these three countries.

```{r ,fig.height=2.5, fig.width=4, fig.align="center"}
statcomp <- dhs_data(indicatorIds = "AN_ANEM_W_ANY", countryIds = dhscc)
statcomp[,.(Indicator, CountryName, SurveyYear, Value, DenominatorWeighted)]

ggplot(statcomp, aes(SurveyYear, Value, col=CountryName)) +
  geom_point() + geom_line()
```

# Analyse DHS microdata

## Identify surveys that include anemia testing

The DHS API provides the facility to filter surveys according to particular characteristics.
We first query the list of survey characteristics and identify the `SurveyCharacteristicID`
that indicates the survey included anemia testing. The first command below queries the API
for the full liest of survey characteristics, and the second uses `grepl()` to search
`SurveyCharacteristicName`s that include the word 'anemia'.

```{r }
surveychar <- dhs_surveyCharacteristics()
surveychar[grepl("anemia", SurveyCharacteristicName, ignore.case=TRUE)]
```

The `SurveyCharacteristicID = 41` indicates that the survey included anemia testing. Next we
query the API to identify the surveys that have this characteristic and were conducted
in our countries of interest.

```{r }
surveys <- dhs_surveys(surveyCharacteristicIds = 41, countryIds = dhscc)
surveys[,.(SurveyId, CountryName, SurveyYear, NumberOfWomen, SurveyNum, FieldworkEnd)]
```

Finally, query the API identify the individual recode (IR) survey datasets for each of these surveys

```{r }
datasets <- dhs_datasets(surveyIds = surveys$SurveyId, fileType = "IR", fileFormat="flat")
datasets[, .(SurveyId, SurveyNum, FileDateLastModified, FileName)]
```

## Download datasets

To get datasets, we establish a `client_dhs` object. The DHS client creates a cache in a specified location to save downloaded datasets and metatdata. The client requires login credentials for accessing datasets from the DHS website, including the email address, password, and project name. Credentials should be stored in a text file with three lines:

```
email=dummy@gmail.com
password=dummypass
project=Dummy Project Name
```

The file path to to this credentials file and the desired location for the cache are passed to the `client_dhs()` function to establish the DHS client.

```{r }
client <- client_dhs(credentials = "credentials",
                     root = file.path(path.expand('~'),'rdhs_cache'))
```

After establishing the client, the function `client$get_datasets()` returns a list of file paths where the desired datasets are saved in the cache. The first time a dataset is accessed, the client will download the dataset from the DHS program website using the supplied credentials. Subsequently, datasets will be simply be located in the cached repository. 

```{r }
datasets$path <- unlist(client$get_datasets(datasets$FileName))
```

## Identify survey variables

Anemia is defined as having a hemoglobin (Hb) <12.0 g/dL for non-pregnant women
or Hb <11.0 g/dL for currently pregnant women^[https://www.measureevaluation.org/prh/rh_indicators/womens-health/womens-nutrition/percent-of-women-of-reproductive-age-with-anemia].
To calculate anemia prevalence from DHS microdata, we must identify the DHS recode
survey variables for hemoglobin measurement and pregnancy status. This could be
done by consulting the DHS recode manual or the .MAP files accompanying survey
datasets. It is convenient though to do this in R by loading the first
individual recode dataset and searching the metadata for the variable
names corresponding to the hemoglobin measurement and pregnancy status.

```{r }
head(client$survey_questions(datasets$FileName[10], "hemoglobin")[,1:2])
```

Variable `v042` records the household selection for hemoglobin testing.
Variable `v455` reports the outcome of hemoglobin measurement and `v456`
the result of altitude adjusted hemoglobin levels.

```{r }
ir <- readRDS(datasets$path[10])
table(as_factor(ir$v042))
table(as_factor(ir$v455))
summary(ir$v456)
```

Variable `v454` reports the current pregnancy status used for determining the
anemia threshold.

```{r }
client$survey_questions(datasets$FileName[1], "currently.*pregnant")[,1:2]
table(as_factor(ir$v454))
```

We also keep a number of other variables related to the survey design and potentially
interesting covariates: country code and phase (`v000`), cluster number (`v001`),
sample weight (`v005`), age (`v012`), region (`v024`), urban/rural residence (`v025`),
and education level (`v106`).

```{r }
vars <- c("SurveyId", "CountryName", "SurveyYear", "v000", "v001", "v005",
          "v012", "v024", "v025", "v106", "v042", "v454", "v455", "v456")
```

## Extract survey data

```{r }
datlst <- list()

for(i in 1:nrow(datasets)){

  if(file.exists(datasets$path[i])){
  
  print(paste(i, datasets$SurveyId[i]))
  ir <- readRDS(datasets$path[i])

  ir$SurveyId <- datasets$SurveyId[i]
  ir$CountryName <- datasets$CountryName[i]
  ir$SurveyYear <- datasets$SurveyYear[i]

  datlst[[datasets$SurveyId[i]]] <- ir[vars]
  }
}
```

We use `rbind_labelled()` to combine datasets with labelled columns. The argument
`labels` describes to combine variable levels for all datasets for `v024` (region)
while providing a consistent set of value labels to be used for `v454` (currently
pregnant) for all datasets.


```{r }
dat <- rbind_labelled(datlst,
                      labels = list(v024 = "concatenate",
                                    v454 = c("no/don't know" = 0L,
                                             "yes" = 1L, "missing" = 9L)))

sapply(dat, is.labelled)
dat$v456 <- zap_labels(dat$v456)
dat <- as_factor(dat)
```

## Data tabulations

It is a good idea to check basic tabulations of the data, especially by
survey to identify and nuances Exploratory analysis of variables

```{r }
with(dat, table(SurveyId, v025, useNA="ifany"))
with(dat, table(SurveyId, v106, useNA="ifany"))
with(dat, table(SurveyId, v454, useNA="ifany"))
with(dat, table(SurveyId, v455, useNA="ifany"))

with(dat, table(v042, v454, useNA="ifany"))
```

## Calculate anemia prevalence
Create indicator variable for 'any anemia'. The threshold depends on pregnancy status.

```{r }
dat$v456[dat$v456 == 999] <- NA
with(dat, table(v455, is.na(v456)))
dat$anemia <- as.integer(dat$v456  < ifelse(dat$v454 == "yes", 110, 120))
dat$anemia_denom <- as.integer(!is.na(dat$anemia))
```

Specify survey design using the `survey` package.

```{r }
dat$w <- dat$v005/1e6
des <- svydesign(~v001+SurveyId, data=dat, weights=~w)

anemia_prev <- svyby(~anemia, ~SurveyId, des, svyciprop, na.rm=TRUE, vartype="ci")
anemia_denom <- svyby(~anemia_denom, ~SurveyId, des, svytotal, na.rm=TRUE)

anemia_prev <- merge(anemia_prev, anemia_denom[c("SurveyId", "anemia_denom")])
res <- statcomp[,.(SurveyId, CountryName, SurveyYear, Value, DenominatorUnweighted, DenominatorWeighted)][anemia_prev, on="SurveyId"]

res$anemia <- 100*res$anemia
res$ci_l <- 100*res$ci_l
res$ci_u <- 100*res$ci_u
res$anemia_denom0 <- round(res$anemia_denom)
```

The table below compares the prevalence of any anemia calcuated from survey microdata
with the estimates from DHS StatCompiler and the weighted denominators for each
calculation. The estimates are identical for most cases. There are some small
differences to be ironed out, which will require looking at the specific coutnries to check
how their stratification was carried out. (We are hoping to bring this feature in once the DHS
program has compiled how sample strata were constructed for all of their studies).

```{r fig.height=2.5, fig.width=4, fig.align="center"}
knitr::kable(res[,.(CountryName, SurveyYear, Value, anemia, ci_l, ci_u, DenominatorWeighted, anemia_denom0)], digits=1)


ggplot(res, aes(x=SurveyYear, y=anemia, ymin=ci_l, ymax=ci_u,
                col=CountryName, fill=CountryName)) +
  geom_ribbon(alpha=0.4, linetype="blank") + geom_point() + geom_line()
```


# Regression analysis: relationship between education and anemia

A key use of the survey microdata are to conduct secondary analysis of pooled data
from several surveys, such as regression analysis. Here we investigate the
relationship between anemia prevalence and education level (`v106`) for women using
logistic regression, adjusting for urban/rural (`v025`) and fixed effects for each
survey.

```{r }
des <- update(des, v106 = relevel(v106, "primary"))
summary(svyglm(anemia ~ SurveyId + v025 + v106, des, family="binomial"))
```

The results suggest that anemia prevalence is lower among women with higher education.
