---
title: "How to use rdhs?"
author: "OJ Watson"
date: "`r Sys.Date()`"
output:
rmarkdown::html_vignette:
css: vignette.css
vignette: >
  %\VignetteIndexEntry{How to use rdhs?}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
  ---
  
  ```{r, echo = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>",
fig.path = "README-"
)
```

## Overview

`rdhs` is a package for management and analysis of [Demographic and Health Survey (DHS)](www.dhsprogram.com) data. This includes functionality to:

1. Access standard indicator data (i.e. [DHS STATcompiler](https://www.statcompiler.com/)) in R via the [DHS API](https://api.dhsprogram.com/).
1. Identify surveys and datasets relevant to a particular analysis.
1. Download survey datasets from the [DHS website](https://dhsprogram.com/data/available-datasets.cfm).
1. Load datasets and associated metadata into R.
1. Extract variables and combining datasets for pooled multi-survey analyses.

This process is described below and should cover most functionality that will be needed for working with these datasets. 

### 0. Installation

You can install rdhs from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("OJWatson/rdhs")

```

### 1. Access standard indicator data via the API

The DHS API has many *endpoints* that can be accessed using anyone of `dhs_<endpoint>()` functions. All exported functions within `rdhs` that start *dhs_* interact with a different with a different endpoint of the [DHS API](https://api.dhsprogram.com/). Their website gives great information about the different search terms and filters that can be used, and we have tried to include all of this within the documentation of each function.

One of those endpoints, `dhs_data()`, interacts with the the publishd set of standard health indicator data calculated by the DHS. This endpoint allows use to retrieve a set of health indicators that have been sample weighted to give country, subnational estimates that can be further refined by education and wealth brackets. To do this we use the `dhs_data()` endpoint, which we can then either search for specific indicators, or by querying for indicators that have been tagged within specific areas.

```{r inddata}
## what are the indicators
indicators <- dhs_indicators()
str(indicators)

```

Each call to a DHS endpoint returns a `data.table` with all the results avaialable by default. The DHS has a unique *IndicatorId* for each of the statistics it calculates. The definition and specific string for each indicator is included within the *IndicatorId* and *Definition* variable:

```{r inddata defs}
# grab the first 5 alphabetically
dplyr::group_by(indicators,IndicatorId) %>% dplyr::summarise(Definition = Definition) %>% head(5)

```

Since there are quite a lot of indicators, it might be easier to first query by tags. The DHS tags their indicators by what areas of demography and health they relate to, e.g. anaemia, literacy, malaria parasitaemia are all specific tags. First let's look at what the tags are, by interacting with the `dhs_tags()` endpoint, before grabbing data that related to malaria parasitaemia in the DRC and Tanzania since 2010:

```{r tags}
# What are the tags
tags <- dhs_tags()

# Let's say we want to view the tags that relate to malaria
tags[grepl("Malaria",tags$TagName),]

# and now let's then grab this data by specifying the countryIds and the survey year starts
data <- dhs_data(tagIds = 36,countryIds = c("CD","TZ"),breakdown="subnational",surveyYearStart = 2010)
str(data)
```

Depending on your analysis this maybe more than enough detail. It is also worth mentioning that this data can also be accessed via [DHS STATcompiler](https://www.statcompiler.com/) if you prefer a click and collect version. However, hopefully one can see that selecting a lot of different indicators for multiple countries and breakdowns should be a lot easier using the `rdhs` API interaction. For example we can very quickly find out the trends in antimalarial use in Africa, and see if perhaps antimalarial prescription has decreased after RDTs were introduced (assumed 2010). 

```{r API call, include=TRUE, message = FALSE, warning = FALSE,fig.height='750px',fig.width='500px'}
# Make an api request
resp <- dhs_data(indicatorIds = "ML_FEVT_C_AML", surveyYearStart = 2010,breakdown = "subnational")

# and let's filter it to 12 countries for space
countries  <- c("Angola","Ghana","Kenya","Liberia",
                "Madagascar","Mali","Malawi","Nigeria",
                "Rwanda","Sierra Leone","Senegal","Tanzania")

# and plot the results
library(ggplot2)
ggplot(resp[resp$CountryName %in% countries,],
       aes(x=SurveyYear,y=Value,colour=CountryName)) +
  geom_point() +
  geom_smooth(method = "glm") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab(resp$Indicator[1]) + 
  facet_wrap(~CountryName,ncol = 6) 

```

If we miss entered a filter query (very possible) rdhs will give you an error that is somewhat nicely presented on the page for you to see way:

```{r API fail, include=TRUE, message = FALSE, error=TRUE, purl= FALSE, warning = FALSE,cache=TRUE}
# Make an api request
resp <- dhs_data(indicatorIds="ML_FEVT_C_AMasfafasfL",
                 surveyYearStart=202231231306,
                 breakdown="subParTyping")

```

### 2. Identify surveys relevant for further analysis

You may, however, wish to do more nuanced analysis than the API allows. The following 4 section detail a very basic example of how to quickly identify, download and extract datasets you are interested in.

Let's say we want to get all the survey data from the Democratic Republic of Congo and Tanzania in the last 5 years (since 2013), which covers the use of rapid diagnostic tests (RDTs) for malaria. To begin we'll interact with the DHS API to identify our datasets.

To start our extraction we'll query the *surveyCharacteristics* endpoint using `dhs_surveyCharacteristics()`:

```{r sc}
## make a call with no arguments
sc <- dhs_surveyCharacteristics()
sc[grepl("Malaria",sc$SurveyCharacteristicName),]

```

There are 87 different survey characteristics, with one specific survey characteristic for Malaria RDTs. We'll use this to then find the surveys that include this characteristic. We can also at this point filter for our desired countries and years. The DHS API allows for countries to be filtered using by their *countryIds*, which is one of the arguments in `dhs_surveys()`. To have a look at what each countries countryId is we can use another of the API endpoints first:

```{r surv}
## what are the countryIds
ids <- dhs_countries(returnFields=c("CountryName","DHS_CountryCode"))
str(ids)

# lets find all the surveys that fit our search criteria
survs <- dhs_surveys(surveyCharacteristicIds = 89,countryIds = c("CD","TZ"),surveyYearStart = 2013)

# and lastly use this to find the datasets we will want to download and let's download the spss (.sav) datasets (have a look in the dhs_datasets documentation for all argument options, and fileformat abbreviations etc.)
datasets <- dhs_datasets(surveyIds = survs$SurveyId, fileFormat = "SV")
str(datasets)
```

we can now use this to download our datasets for further analysis. 

### 3. Download survey datasets

We can now go ahead and download our datasets. To do this we need to first create a `client`. The client is an R6 class (similar to R's built in reference classes and make caching survey and API queries more reproducible) and will be used to log in to your DHS account, download datasets for you, and help query those datasets for the question you are interested in. The client will also cache all of these processes, which really helps increase the reproducibility of your analysis. 

To create our client we use the `client()` function and you need to specify your log in credentials for the DHS website. This is best provided, for security reasons, by giving a path to a file that contains your email, password and project title that you used when setting up your account with the DHS website. This should take the form of a file path that contains 3 lines, something like this:

* email="dummy\@gmail.com"
* password="dummypass"
* project="Dummy Project"

It also takes an argument for its `root`, which is the directory path where the client and associated caches will be stored. If left bank, a suitable directory will be created within your user cache directory for your operating system.

```{r client}
## create a client
client <- dhs_client(credentials = "credentials")
client

```

Before we use our client to download our datasets, it is worth mentioning that the client can be passed as an argument to any of the API functions we have just seen. This will then cache the results for you, so that if you are working remotely or without a good internet connection you can still return your previous API requests:

```{r client_api_cache}
# before it's cached we provide the client so the results is cached within our client
s <- dhs_surveys(client = client)

# with it cached it will be returned much quicker with the client argument
microbenchmark::microbenchmark(dhs_surveys(client = client),times = 1)

# without it cached again for comparison
microbenchmark::microbenchmark(dhs_surveys(),times = 1)

```

Now back to our dataset downloads. If we have a look back at our datasets object, we'll see there are 19 datasets listed. However, not all of them will be relevant to our malaria RDT questions. One approach is to head to the DHS website and have a look at the [DHS Recodes](https://dhsprogram.com/publications/publication-dhsg4-dhs-questionnaires-and-manuals.cfm), and look at the recodes that relate to the surveys. The other alternative is to download all the surveys and then query the variables within them. This is what we'll demonstrate here as it also demonstrates more of the package's functionality:

So first we will download all these datasets:

```{r download, message=FALSE}
# download datasets
downloads <- client$download_datasets(datasets$FileName)

str(downloads)
```

The function returns a list with a file path to where the downloaded datasets have been saved to. 

### 4. Load datasets and associated metadata into R.

We can now examine what it is we have actually downloadsed, by reading in one of these datasets:

```{r read a dataset}
# read in our dataset
cdpr <- readRDS(downloads$CDPR61SV)
```

The dataset returned here is a list that contains the *dataset* but also a data.frame called *variable_names*. This contains all the survey questions within the dataset, and what their survey variable is:

```{r probe dataset}
# let's look at the variable_names
head(cdpr$variable_names)

# and then the dataset
class(cdpr$dataset)
str(cdpr$dataset$hv024)
```

This is the default behaviour for the client function `download_datasets` - it will download the datasets for you, and then by default save them in your client's root directory and then unzip them and read them in for you, and save the resultant data.frame as a .rds object within the client's root directory. You can control this behaviour using the `download_option` argument as such:

* `client$download_datasets(download_option = "zip")` - Just the downloaded zip will be saved
* `client$download_datasets(download_option = "rds")` - Just the read in rds will be saved
* `client$download_datasets(download_option = "both")` - The zip is downloaded and saved as well as the read in rds

The other main reason for reading the dataset in straight away as the default option is that the created table of all the survey variables and their definitions is cached then and there, which then allows us to quickly query for particular search terms or survey variables:

```{r questions}
# rapid diagnostic test search
questions <- client$survey_questions(datasets$FileName,search_terms = "malaria rapid test")

table(questions$dataset_filename)
```

What we see from the questions is that the question "Result of malaria rapid test" appears in a few different datasets. This is because the household member recode datasets (CDPR61SV, TZPR7HSV) stores information about the children in a household, with one row per child, whereas the household recode (CDHR61SV, TZHR7HSV) stores information about the household, and thus flattens the information from each child into different subvariables (hml35$01/02 etc). As such it is easier to extract this information from the household member recodes. 

### 5. Extract variables and combining datasets for pooled multi-survey analyses.

To extract our data we pass our questions object to the client function `extract`, which will create a list with each dataset and its extracted data as a `data.frame`. We also have the option to add any geographic data available, which will download the geographic data files for you and add this data to you resultant extract:

```{r extract_questions}
# let's just use the PR files thus
datasets <- dhs_datasets(surveyIds = survs$SurveyId, fileFormat = "SV",fileType = "PR")
downloads <- client$download_datasets(datasets$FileName)

# and grab the questions from this again along with also questions detailing the province
questions <- client$survey_questions(datasets$FileName,search_terms = c("malaria rapid test"))

# and now extract the data
extract <- client$extract(questions,add_geo = TRUE)

# what does our extract look like
str(extract)
```

The resultant extract is a list, with a new element for each different dataset that you have extracted. The responses from the dataset are by default stored as a *labelled* class from the [haven package](https://github.com/tidyverse/haven). This class preserves the original semantics and can easily be coerced to factors with `haven::as_factor()`. Special missing values are also preserved. For more info on the *labelled* class have a look at their github.

We can also query our datasets for the survey question variables. In the example above the survey question was *Result of malaria rapid test* and the variable was *hml35*. So if you knew the survey variables that you wanted (either by looking at the Recode file or by looking through the *variable_names* included in the datasets) then we could search against these. So let's grab the regions using *hv024* using the client function `survey_variables()`:

```{r extract_variables}
# and grab the questions from this now utilising the survey variables
questions <- client$survey_variables(datasets$FileName,variables = c("hv024","hml35"))

# and now extract the data
extract2 <- client$extract(questions,add_geo = TRUE)

# quick check
head(extract2$CDPR61SV)
head(extract2$TZPR7HSV)

```

We can now combine our two dataframes for further analysis using the `rdhs` package function `rbind_labelled()`. This function works specifically with our lists of labelled data.frames:

```{r rbind_labelled}
# first let's bind our first extraction, without the hv024
extract_bound <- rbind_labelled(extract)

head(extract_bound)

# now let's try our second extraction
extract2_bound <- rbind_labelled(extract2)

```

This hasn't quite done what we might want in the second instance. The *hv024* variable stores the regions for these 2 countries, which will not be the same and thus the labels will be different between the two of them. Without specifying any additional arguments `rbind_labelled()` will simply use the first data.frames labelling as the default, which will mean that some of the Tanzanian provinces will have been encoded as DRC provinces - not good! (This is a similar problem in nature to say trying to add new character strings to a factored data.frame).

There are a few work arounds. Firstly, we can specify a *labels* argument to the function which will detail how we should handle different variables. *labels* is a names list that specifies how to handle each variable. If we simply want to keep all the labels then we us the string "concatenate":

```{r concatenate}
# lets try concatenating the hv024
better_bound <- rbind_labelled(extract2,labels = list("hv024"="concatenate"))

head(better_bound$hv024)

```

We could also specify new labels for a variable. For example, imagine the two datasets encoded their RDT responses differently, with the first one as `c("No","Yes")` and the other as `c("Negative","Positive")`. These would be for our purposes the same response, and so we could either leave it and all our results would use the `c("No","Yes")` labelling. But we may want to use the latter as it's more informative/correct, or we may want to be crystal clear and use `c("NegativeTest","PositiveTest")`. we can do that like this:

```{r concatenate and new label}
# lets try concatenating the hv024 and providing new labels
better_bound <- rbind_labelled(extract2,
                               labels = list("hv024"="concatenate",
                                             "hml35"=c("NegativeTest"=0,"PositiveTest"=1)))

# and our new label
head(better_bound$hml35)
```

The other option is to not use the labelled class at all. We can control this when we download our datasets, using the argument `reformat=TRUE`. This will ensure that no factors or labels are used and it is just the raw data:

```{r reformat}
# grab the questions but specifying the reformat argument
questions <- client$survey_variables(datasets$FileName,variables = c("hv024","hml35"),
                                     reformat=TRUE)

# and now extract the data
extract3 <- client$extract(questions,add_geo = TRUE)

# group our results
bound_no_labels <- rbind_labelled(extract3)

# what does our hv024 look like now
table(bound_no_labels$hv024)

```

The *hv024* column in now just characters, which is possibly the best option depending on your downstream analysis/preferences. It's for this reason that the geographic data that is added is never turned into factors or labels.  

Lastly, we can now use our extract dataset to carry out some regression analysis, to investigate the relationship between malaria prevalence and the quality of wall materials. To do this we will need to first grab the sample weights and stratification from the surveys, along with the extra variables and we will then check the RDT prevalence calculated using the raw data versus the API:

```{r regression analysis}
# grab the additional variable hv023 and hv024 which have the strata and weights respectively
questions <- client$survey_variables(datasets$FileName,variables = c("hv005","hv023","hv024","hv025","hv214","hml20","hml35"))
extraction <- client$extract(questions,TRUE)

# now concatenate the provinces as before and remove missing responses
dat <- rbind_labelled(extraction,labels=list("hv024"="concatenate"))
dat <- dat[-which(dat$hml35==9),] # remove missing responses

# create a denominator response for hml35
dat$hml35denom <- as.integer(!is.na(dat$hml35))
dat$bricks <- dat$hv214 %in% c(31,32,33,34,35)
dat$net <- as.logical(dat$hml20)
# specif the strata and sample weights
dat$strata <- paste0(dat$hv023,dat$DATASET)
dat$hv005 <- dat$hv005/1e6

# construct a survey design using the survey pacakge
library(survey)

# construct the sample design and calculate the mean and totals
des <-  survey::svydesign(~CLUSTER+DATASET,data=dat,weight=~hv005)
results <- cbind(survey::svyby(~hml35,by=~DHSREGNA, des, survey::svyciprop,na.rm=TRUE),
                 survey::svyby(~hml35denom,by=~DHSREGNA, des, survey::svytotal))

# grab the same data from the API
dhs_api_data <- dhs_data(countryIds = c("CD","TZ"),indicatorIds = "ML_PMAL_C_RDT",breakdown = "subnational",surveyYearStart = 2013)
dhs_api_data <- cbind(dhs_api_data$Value,dhs_api_data$DenominatorWeighted,dhs_api_data$CharacteristicLabel)
api <- dhs_api_data[!grepl("\\.\\.",dhs_api_data[,3]),] # remove subregions included in Tanzania
api <- api[order(api[,3]),]

# bind the results and remove duplicate Region Columns
comparison <- cbind(results[,-c(3,4,6)],api[])
names(comparison) <- c("Region","Survey_RDT_Prev","Survey_RDT_Denom","API_RDT_Prev","API_RDT_Denom","API_Regions")
comparison
```

It's a little off, which will be due to ...

And lastly we will construct a logistic regression to investigat the relationship between a positive malaria RDT and whether the main walls of an individual's house were made of bricks or similar, while adjusting for urban/rural (`v025`) and fixed effects for each survey.

```{r altitutde glm}
# contsruct our glm using svyglm and specify quasibinomial to handle the na in hml35
summary(svyglm(hml35 ~ DATASET + hv025 + net + bricks, des, family="binomial"))

```

What we can see is that a significant negative gradient was associated with walls being made of bricks or similarly good materials. What is also interesting is that whether the individual slept under a long lasting insecticial net (*hml20* that we converted to net) was not significant.  

---

## Summary and further thoughts

Hopefully the above tutorial has shown how the `rdhs` package can facilitate both querying the DHS API and hopefully make downloadeding and interacting with the raw datasets a smoother, more reprodcuable process. It is worth bearing in mind thogh, that creating a harmonised dataset is not always as easy as the example above - a lot of the time survey variables differ across years and surveys, wich is hopefully when the `survey_questions` functionality will make it easier to first filter down to those that include the relevant questions before having to decide which survey questions are valid. 

Any suggestions or comments/corrections/errors/ideas please let me know either in the issues or send me an email at "o.watson15@imperial.ac.uk". And if there is any further functionality that you think you wouldbe useful, then also let me know. :)
