---
title: "Downloading Shape Files for DHS Surveys"
author: "OJ Watson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interacting with geojson API results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

`rdhs` can be used to download shape files associated with the DHS surveys. For
example, we can link API responses to their shape files and plot the subnational
estimates using the spatial pacakge `sp`. To do this, we use the new function
`download_boundaries` from `rdhs` to download shape files. 

```{r normal, warnings = FALSE, message = FALSE, fig.width=6, fig.height=6}

# load our package
library(rdhs)

# make request
d <- dhs_data(countryIds = "SN",
              indicatorIds = "FE_FRTR_W_A15",
              surveyYearStart = 2014,
              breakdown = "subnational")

# get our related spatial data frame object
sp <- download_boundaries(surveyId = d$SurveyId[1])

# match our values to the regions
m <- d$Value[match(sp$sdr_subnational_boundaries$REG_ID, d$RegionId)]
sp$sdr_subnational_boundaries@data$Value <- m

# Use sp to plot
sp::spplot(sp$sdr_subnational_boundaries, "Value", main = d$Indicator[1])

```

Or we can use `sf` for our plotting, which offers more user-friendly plotting
options. 

```{r sf, warnings = FALSE, message = FALSE, fig.width=6, fig.height=6}

# make request
d <- dhs_data(countryIds = "SN",
              indicatorIds = "FE_FRTR_W_A15",
              surveyYearStart = 2017,
              breakdown = "subnational")

# get our related spatial data frame object
sp <- download_boundaries(surveyId = d$SurveyId[1], method = "sf")

# match our values to the regions
m <- d$Value[match(sp$sdr_subnational_boundaries$REG_ID, d$RegionId)]
sp$sdr_subnational_boundaries$Value <- m

# Use ggplot and geom_sf to plot
library(ggplot2)
ggplot(sp$sdr_subnational_boundaries) + 
  geom_sf(aes(fill = Value)) + 
  ggtitle(d$Indicator[1])

```
